Epilepsy example

### 2022-11-02 src/epil/epil.Rmd
Update epil writing and add Fig 3. (a) from INLA 2009 paper

Section 5.2. of @rue2009approximate demonstrates the application of \texttt{R-INLA} to a generalised linear mixed model.
The data, previously included as an example in the OpenBUGs manual, concern the results of a clinical trial for a new epilepsy drug.
Patients $i = 1, \ldots, 59$ are each either assigned the new drug $\texttt{Trt}_i = 1$ or placebo $\texttt{Trt}_i = 0$.
Each patient makes four visits the clinic $j = 1, \ldots, 4$.
The observations $y_{ij}$ give the number of seizures of the $i$th person in the two weeks preceeding their $j$th visit to the clinic.
The covariates are age $\texttt{Age}_i$, baseline seizure counts $\texttt{Base}_i$ and an indicator for the final clinic visit $\texttt{V}_4$, which are all centered.
The observations are modelling using a Poisson distribution $y_{ij} \sim \text{Poisson}(e^{\eta_{ij}})$ with linear predictor 
\begin{align*}
  \eta_{ij}       &= \beta_0 + \beta_\texttt{Base} \log(\texttt{Baseline}_j / 4) + \beta_\texttt{Trt} \texttt{Trt}_i +
                     \beta_{\texttt{Trt} \times \texttt{Base}} \texttt{Trt}_i \times \log(\texttt{Baseline}_j / 4) \\ 
                  &+ \beta_\texttt{Age} \log(\texttt{Age}_i) + \beta_{\texttt{V}_4} {\texttt{V}_4}_j +
                     \epsilon_i + \nu_{ij}, \quad i=1:59, \quad j=1:4,
\end{align*}
where the prior on each of the regression parameters is $\mathcal{N}(0, 100^2)$.
The random effects are IID $\epsilon_i \sim \mathcal{N}(0, 1/\tau_\epsilon)$ and $\nu_{ij} \sim \mathcal{N}(0, 1/\tau_\nu)$ with precision priors $\tau_\epsilon, \tau_\nu \sim \Gamma(0.001, 0.001)$.

@@

Start by centering the covariates:

@@

Implement the model based upon a response variable of length $N \times J$ and a model matrix with $K$ predictors (including the intercept term):

@@

For the individual specific random effect $\epsilon_i$ we use a transformation matrix $\bm{E}$ which repeats elements of vector of length `N` each `J` times, for example:

```{r}
make_epsilon_matrix <- function(N, J) {
  t(outer(1:N, 1:(N * J), function(r, c) as.numeric((J*(r - 1) < c) & (c <= J*r))))
}

E <- make_epsilon_matrix(N = 3, J = 2)
E
t(E %*% c(1, 2, 3)) # Same as rep(1:3, 2)
```

@@

@@

Hamiltonian Monte Carlo via the No-U-Turn algorithm, with one Markov chain of `iter = 1000` samples, and `warmup = 100` burn-in:

@@

fit1 <- stan(
  "epil.stan",
  data = dat,
  chains = 1,
  warmup = 100,
  iter = 1000,
  control = list(adapt_delta = 0.95)
)
```

@@

The variable `Epil$rand` is gives the row number for each entry and the variable `Epil$Ind` gives the patient number.
These variables can be used to define the random effects $\nu_{ij}$ and $\epsilon_i$ in INLA as follows.
The usual R formula notation (e.g. `y ~ x1 + x2`) is used for the rest of the linear predictor.

```{r}
formula <- y ~ 1 + CTrt + ClBase4 + CV4 + ClAge + CBT +
  f(rand, model = "iid", hyper = tau_prior) +  # Nu random effect
  f(Ind,  model = "iid", hyper = tau_prior)    # Epsilon random effect
```

We can use various settings for the strategy, which controls how the latent field posterior marginals are approximated:

```{r}
epil_inla <- function(strat) {
  inla(formula,
       control.fixed = list(mean = 0, prec = 1/100^2), # Beta prior
       family = "poisson",
       data = Epil,
       control.inla = list(strategy = strat),
       control.predictor = list(compute = TRUE))
}

@@

We integrate `random` out with a Laplace approximation:

@@

For completeness, we check here that the objective functions as implemented in Stan and `TMB` are identical.
To obtain the `TMB` negative log-likelihood we call `MakeADFun` as before, but now do not specify any parameters to be integrated out.
For Stan, we create an empty model, then use `rstan::log_prob`:

@@

data.frame(
  "TMB" = tmb_evals,
  "Stan" = stan_evals
) %>%
  mutate(Difference = TMB - Stan)
```

@@

[`glmmTMB`](https://glmmtmb.github.io/glmmTMB/) is an R package written by Ben Bolker and collaborators which allows fitting generalised linear mixed models in `TMB` without writing the C++ code.
For example, rather than writing `epil.cpp` manually, we could have called the following:

@@

This likely isn't the same model, in terms of priors and so on, but it may be close.

@@

```{r}
fit7 <- tmbstan(obj = obj, chains = 1)
```

@@

[`aghq`](https://arxiv.org/pdf/2101.04468.pdf) is an R package written by Alex Stringer which implements adaptive Gauss-Hermite quadrature.
One (strange!) approach is to use `glmmTMB` to get the `TMB` template which can be passed the `aghq`.

@@

```{r results='hide'}
fit8 <- aghq::marginal_laplace_tmb(
  obj,
  k = 2,
  startingvalue = c(param$beta, param$l_tau_epsilon, param$l_tau_nu)
)

@@

beta_i <- function(i) { c(paste0("beta_", i), paste0("sd(beta_", i, ")")) }
rownames(df) <- c(sapply(0:5, beta_i))
colnames(df) <- c("Stan", "INLA_G", "INLA_SL", "INLA_L", "TMB", "glmmTMB", "tmbstan", "aghq", "type")

kableExtra::kable(df)

@@

@@

  ggplot(marginals, aes(x = x, y = y, group = method, col = method)) +
    geom_line() +
    theme_minimal()
}

>>>>

## Model description

Section 5.2. of @rue2009approximate demonstrates the application of `R-INLA` to a generalised linear mixed model.
The data, previously included as an example in the OpenBUGs manual, concern the results of a clinical trial for a new epilepsy drug.
Patients $i = 1, \ldots, 59$ are each either assigned the new drug $\texttt{Trt}_i = 1$ or placebo $\texttt{Trt}_i = 0$.
Each patient makes four visits the clinic $j = 1, \ldots, 4$.
The observations $y_{ij}$ give the number of seizures of the $i$th person in the two weeks preceeding their $j$th visit to the clinic.
The covariates are age $\texttt{Age}_i$, baseline seizure counts $\texttt{Base}_i$ and an indicator for the final clinic visit $\texttt{V}_4$, which are all centered.
The observations are modelling using a Poisson distribution $y_{ij} \sim \text{Poisson}(e^{\eta_{ij}})$ with linear predictor 
\begin{align*}
  \eta_{ij}       &= \beta_0 + \beta_\texttt{Base} \log(\texttt{Baseline}_j / 4) + \beta_\texttt{Trt} \texttt{Trt}_i +
                     \beta_{\texttt{Trt} \times \texttt{Base}} \texttt{Trt}_i \times \log(\texttt{Baseline}_j / 4) \\ 
                  &+ \beta_\texttt{Age} \log(\texttt{Age}_i) + \beta_{\texttt{V}_4} {\texttt{V}_4}_j +
                     \epsilon_i + \nu_{ij}, \quad i=1:59, \quad j=1:4,
\end{align*}
where the prior on each of the regression parameters is $\mathcal{N}(0, 100^2)$.
The random effects are IID $\epsilon_i \sim \mathcal{N}(0, 1/\tau_\epsilon)$ and $\nu_{ij} \sim \mathcal{N}(0, 1/\tau_\nu)$ with precision priors $\tau_\epsilon, \tau_\nu \sim \Gamma(0.001, 0.001)$.

## Data preparation

The data is available within `R-INLA` as follows:

@@

We modify `Epil` by centering all of the covariates:

@@

Implement the model based upon a response variable of length $N \times J$, where $N$ is the number of patients and $J$ is the number of visits to the clinic, and a model matrix with $K$ predictors (including the intercept term):

@@

For the individual specific random effect $\epsilon_i$ we use a transformation matrix `E` which repeats elements of vector of length `N` each `J` times.

```{r}
make_epsilon_matrix <- function(N, J) {
  t(outer(1:N, 1:(N * J), function(r, c) as.numeric((J*(r - 1) < c) & (c <= J*r))))
}
```

For example, with $N = 3$ and $J = 2$:

```{r}
E <- make_epsilon_matrix(N = 3, J = 2)
E
t(E %*% c(1, 2, 3)) # Same as rep(1:3, 2)
```

@@

We consider a range of inference methods, described below.

@@

First, Hamiltonian Monte Carlo via the No-U-Turn algorithm, with four Markov chains of `iter = 1000` samples, and `warmup = 100` burn-in:

@@

fit1 <- stan(
  "epil.stan",
  data = dat,
  chains = 4,
  warmup = 100,
  iter = 1000,
  control = list(adapt_delta = 0.95)
)
```

@@

The variable `Epil$rand` is gives the row number for each entry and the variable `Epil$Ind` gives the patient number.
These variables can be used to define the IID random effects $\nu_{ij}$ and $\epsilon_i$ in `R-INLA` as follows.
The usual R formula notation (e.g. `y ~ x1 + x2`) is used for the rest of the linear predictor.

```{r}
formula <- y ~ 1 + CTrt + ClBase4 + CV4 + ClAge + CBT +
  f(rand, model = "iid", hyper = tau_prior) +  #' Nu random effect
  f(Ind,  model = "iid", hyper = tau_prior)    #' Epsilon random effect
```

We can use various settings for the strategy, which controls the approximation used for the latent field posterior marginals:

```{r}
epil_inla <- function(strat) {
  inla(formula,
       control.fixed = list(mean = 0, prec = 1 / 100^2), #' Beta prior
       family = "poisson",
       data = Epil,
       control.inla = list(strategy = strat),
       control.predictor = list(compute = TRUE))
}

@@

We integrate `random` out, which here corresponds to `c("epsilon", "nu")`, with a Laplace approximation:

@@

### Check `TMB` objective is the same as Stan

For completeness, we check here that the objective functions as implemented in `TMB` and Stan are identical.
To obtain the `TMB` negative log-likelihood we call `MakeADFun` as before, but now do not specify any parameters to be integrated out.
For Stan, we create an empty model, then use `rstan::log_prob`:

@@

data.frame(
  "TMB" = tmb_evals,
  "Stan" = stan_evals
) %>%
  mutate(Difference = TMB - Stan) %>%
  pull(Difference)
```

@@

[`glmmTMB`](https://glmmtmb.github.io/glmmTMB/) is an R package written by Ben Bolker and collaborators which allows fitting generalised linear mixed models in `TMB` without writing the C++ code manually as we have in `epil.cpp`.
For example, rather than writing `epil.cpp`, we could have called the following:

@@

This won't be exactly the same model, in terms of priors and so on, but it may be close.

@@

```{r}
fit7 <- tmbstan(obj = obj, chains = 4)
```

@@

[`aghq`](https://arxiv.org/pdf/2101.04468.pdf) is an R package written by Alex Stringer which implements adaptive Gauss-Hermite quadrature.
It also works smoothly with `TMB` templates.
One (odd!) approach would be to use `glmmTMB` to get the `TMB` template which can be passed the `aghq`.

@@

```{r results='hide'}
fit8 <- aghq::marginal_laplace_tmb(
  obj,
  k = 3,
  startingvalue = c(param$beta, param$l_tau_epsilon, param$l_tau_nu)
)

@@

beta_i <- function(i) { c(paste0("beta_", i), paste0("sd(beta_", i, ")")) }
rownames(df) <- c(sapply(0:5, beta_i))
colnames(df) <- c("Stan", "R-INLA_G", "R-INLA_SL", "R-INLA_L", "TMB", "glmmTMB", "tmbstan", "aghq", "type")

df %>%
  select(-type) %>%
  kableExtra::kable()

@@

## Focus on $\beta_0$

```{r}
plot_pdf <- ggplot(data.frame(x = rstan::extract(fit1, pars = "beta[1]")[[1]]), aes(x = x)) +
    geom_histogram(aes(y = ..density..), alpha = 0.8, fill = cbpalette[7], bins = 60) +
    theme_minimal() +
    labs(x = "beta[1]", y = "Posterior PDF")

beta0_inla_marginals <- bind_rows(
  data.frame(fit2$marginals.fixed$`(Intercept)`) %>%
    mutate(method = "G"),
  data.frame(fit3$marginals.fixed$`(Intercept)`) %>%
    mutate(method = "SL"),
  data.frame(fit4$marginals.fixed$`(Intercept)`) %>%
    mutate(method = "L")
)

plot_pdf +
  geom_line(data = beta0_inla_marginals, aes(x = x, y = y, col = method)) +
  scale_color_manual(values = cbpalette)
```

@@

  ggplot(marginals, aes(x = x, y = y, group = method, col = method)) +
    geom_line() +
    scale_color_manual(values = cbpalette)
    theme_minimal()
}
